---
title: "PLOS_BERGMANN_et_al"
author: "C.Bergmann"
date: "2025-09-05"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

Content: This R Markdown document includes the majority of 'Age-related increase in neurofilament light in blood among mammalian species' analyses - German Center for Neurodegenerative Diseases (DZNE) e.V. Tübingen & Hertie Institute for Clinical Brain Research - Department of Cellular Neuroscience - AG Jucker


# **Installing & loading libraries**
## Installing libraries
```{r, message = FALSE}
# List of packages
packages <- c(
  "readxl", "dplyr", "writexl", "ggplot2", "openxlsx", "data.table",
  "vroom", "readr", "tidyr", "survival", "survminer", "finalfit",
  "lme4", "corrplot", "lmerTest", "psych", "mnormt", "DescTools",
  "patchwork", "tidyverse", "scales", "ggthemes", "ggcorrplot",
  "OlinkAnalyze", "cowplot", "ggpubr", "tibble", "purrr", "pheatmap",
  "optimx", "bestNormalize", "car", "MASS", "nortest", "broom", "dunn.test",
  "mgcv", "ggbreak", "ggtext")

for (pkg in packages) {
  if (!require(pkg, character.only = TRUE)) {
    message(paste0("📦 ", pkg, " is installed…"))
    install.packages(pkg, dependencies = TRUE)
  }
  message(paste0("✅ ", pkg, "is loaded"))
  library(pkg, character.only = TRUE)
}

message("All packages are ready!")

```

# **FIGURE 1**
```{r}
##### LOAD, MERGE, PREPARE DATA #####
# Upload Excel file
NfL_Species <- read_excel("/Users/carinabergmann_1/Documents/01-Carina_PhD/28-NfL-Species-Paper/19-DATA/00-Bergmann_et_al_Original.xlsx")
  
NfL_Species$Lower_age_limit <- 0.05*NfL_Species$`Mean life expectancy`
NfL_Species$Upper_age_limit <- 0.5*NfL_Species$`Mean life expectancy`

NfL_Species$Fig_1_included <- ifelse(
  # --> No Age information known --> Fig_1_included = 2
  is.na(NfL_Species$Age_years), 2,
  # --> If age is above lower age limit --> Fig_1_included = 1, otherwise NA 
  ifelse(NfL_Species$Age_years >= NfL_Species$Lower_age_limit, 1, NA)) 

NfL_Species %>%
  filter(Figures %in% c(1, NA)) %>%   # if sample shown in figures -> 1 if not -> NA
  count(Figures, Fig_1_included)


# Figures = 1 & Fig_1_included = 1 --> in figures & age above 0.05%
# Figures = 1 & Fig_1_included = 2 --> in Figure 4 (no age known)
# Figures = 1 & Fig_1_included = NA --> in Figure 4 (very young)
# Figures = NA & Fig_1_included = 1 --> NOT in figures although age above 0.05%
# Figures = NA & Fig_1_included = 2 --> NOT in figures & no age known
# Figures = NA & Fig_1_included = NA --> NOT in figures, too young


NfL_Species$Sup_2_included <- ifelse(NfL_Species$Age_years >= NfL_Species$Lower_age_limit & NfL_Species$Age_years <= NfL_Species$Upper_age_limit, 1, NA)
NfL_Species$Version <- as.factor(NfL_Species$Version)

# Filter data for species - Figure 1
NfL_Species_Figure1 <- NfL_Species %>%
  filter(
    Species %in% c("Mouse", "Cat", "Dog", "Human", "Horse"),
    Fig_1_included == 1)

# FILTER SPECIES 
NfL_mouse <- NfL_Species_Figure1 %>%
  filter(Species == "Mouse")
NfL_cat <- NfL_Species_Figure1 %>%
  filter(Species == "Cat" & Age_years > 1)  
NfL_dog <- NfL_Species_Figure1 %>%
  filter(Species == "Dog" & Age_years > 1) 
NfL_human <- NfL_Species_Figure1 %>%
  filter(Species == "Human")
NfL_horse <- NfL_Species_Figure1 %>%
  filter(Species == "Horse")

###### 1A ######
mouse_Group_1 <- NfL_mouse %>%
  filter(Age_years <= 1)
mouse_Group_2 <- NfL_mouse %>%
  filter(Age_years > 1 & Age_years <= 2)
mouse_Group_3 <- NfL_mouse %>%
  filter(Age_years > 2)

cat_Group_1 <- NfL_cat %>%
  filter(Age_years <= 7 & Age_years > 1)
cat_Group_2 <- NfL_cat %>%
  filter(Age_years > 7 & Age_years <= 14)
cat_Group_3 <- NfL_cat %>%
    filter(Age_years > 14)

dog_Group_1 <- NfL_dog %>%
  filter(Age_years <= 6 & Age_years > 1)
dog_Group_2 <- NfL_dog %>%
  filter(Age_years > 6 & Age_years <= 12)
dog_Group_3 <- NfL_dog %>%
  filter(Age_years > 12)

human_Group_1 <- NfL_human %>%
  filter(Age_years <= 40)
human_Group_2 <- NfL_human %>%
  filter(Age_years > 40 & Age_years <= 80)
human_Group_3 <- NfL_human %>%
  filter(Age_years > 80)

horse_Group_1 <- NfL_horse %>%
  filter(Age_years <= 12.5)
horse_Group_2 <- NfL_horse %>%
  filter(Age_years > 12.5 & Age_years <= 25)
horse_Group_3 <- NfL_horse %>%
  filter(Age_years > 25)

# Bind rows and name groups
Figure_1_A <- bind_rows(
  mouse_Group_1 %>% mutate (Group = "Mouse <= 1 year", Set = "Set1"),
  mouse_Group_2 %>% mutate (Group = "Mouse > 1 - 2 years", Set = "Set1"),
  mouse_Group_3 %>% mutate (Group = "Mouse > 2 years", Set = "Set1"),
  cat_Group_1 %>% mutate (Group = "Cat <= 7 years", Set = "Set2"),
  cat_Group_2 %>% mutate (Group = "Cat > 7 - 14 years", Set = "Set2"),
  cat_Group_3 %>% mutate (Group = "Cat > 14 years", Set = "Set2"),
  dog_Group_1 %>% mutate (Group = "Dog <= 6 years", Set = "Set3"),
  dog_Group_2 %>% mutate (Group = "Dog > 6 - 12 years", Set = "Set3"),
  dog_Group_3 %>% mutate (Group = "Dog > 12 years", Set = "Set3"),
  human_Group_1 %>% mutate (Group = "Human 20 - 40 years", Set = "Set4"),
  human_Group_2 %>% mutate (Group = "Human > 40 - 80 years", Set = "Set4"),
  human_Group_3 %>% mutate (Group = "Human > 80 years", Set = "Set4"),
  horse_Group_1 %>% mutate (Group = "Horse <= 12.5 years", Set = "Set5"),
  horse_Group_2 %>% mutate (Group = "Horse > 12.5 - 25 years", Set = "Set5"),
  horse_Group_3 %>% mutate (Group = "Horse > 25 years", Set = "Set5"))


```


### *Figure 1 A* 
```{r}
###### FINAL PLOT CREATED WITH GRAPHPAD PRISM ######

# Median and 95%CI
summary_Figure_1_A <- Figure_1_A %>%
  group_by(Group) %>%
  summarise(
    median = median(NfL),
    sd = sd(NfL),
    n = n(),
    se = sd/sqrt(n),
    ci = qt(0.975, df = n-1) * se,
    .groups = "drop")
# Order
Figure_1_A$Group <- factor(Figure_1_A$Group, levels = unique(Figure_1_A$Group))
# Plot
ggplot(Figure_1_A, aes(x = Group, y = NfL)) +
  geom_point(size = 1, color = "black", alpha = 0.6, 
           position = position_jitter(width = 0.1), inherit.aes = FALSE,
           aes(x = Group, y = NfL)) +
  stat_summary(fun.data = mean_ci, geom = "errorbar", width = 0.02, color = "darkred") +
  stat_summary(fun = median, geom = "point", 
             shape = 95, size = 6, color = "red") + 
  geom_hline(yintercept = 0.696, linetype = "dashed", color = "darkblue") +
  geom_hline(yintercept = 1.38, linetype = "dashed", color = "lightblue") +  
  labs(y = "NfL [pg/ml]", x = "") +
  scale_y_log10(
    limits = c(0.01, 10000),
    breaks = c(0.01, 0.1, 1, 10, 100, 1000, 10000),
    labels = scales::comma) +
  geom_vline(xintercept = c(3.5, 6.5, 9.5, 12.5), linetype = "dashed") +
  theme_classic2 (base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#######



# 3WAY ANOVA including sex
ANOVA_1A <- Figure_1_A %>%
  filter (Gender %in% c("m","f"))
Table_1A <- ANOVA_1A %>%
  group_by (Species, Group) %>%
  summarise (
    Number = n(),
    Females = sum(Gender == "f", na.rm = TRUE),
    Males = sum(Gender == "m", na.rm = TRUE),
    Unknown_sex = sum (Gender == "u", na.rm = TRUE))
   
# Table_1A

ANOVA_1A <- ANOVA_1A %>%
  mutate (
    Age_group = case_when(
      Group %in% c("Mouse <= 1 year", "Cat <= 7 years", "Dog <= 6 years", "Human 20 - 40 years","Horse <= 12.5 years")  ~ "Young",
      Group %in% c("Mouse > 1 - 2 years", "Cat > 7 - 14 years", "Dog > 6 - 12 years", "Human > 40 - 80 years","Horse > 12.5 - 25 years")  ~ "Adult",
      Group %in% c("Mouse > 2 years", "Cat > 14 years", "Dog > 12 years", "Human > 80 years", "Horse > 25 years") ~ "Aged",
      TRUE                           ~ NA_character_))

options(contrasts = c("contr.sum", "contr.poly"))  # wichtig für Type III
fit <- lm(NfL_log ~ Species * Age_group * Gender, data = ANOVA_1A)
# Type III
Anova(fit, type = 3)

```

### *Figure 1 B* 
```{r}
Figure_1B <- subset(NfL_Species_Figure1, !(Species %in% c("Cat", "Dog") & Age_years <= 1))


Figure_1B_plot <- ggplot(Figure_1B, aes(x = Age_years, y = NfL, color = Species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(aes(group = Species), method = "gam", formula = y ~ s(x, bs = "cs", k = 5), se = FALSE) +
  scale_x_log10(
    limits = c(0.1, 110),
    breaks = c(0.1, 1, 10, 100),
    labels = c(0.1, 1, 10, 100)) +
  scale_y_log10(
    limits = c(0.1, 10000),
    breaks = c(0.1, 1, 10, 100, 1000, 10000),
    labels = c(0.1, 1, 10, 100, 1000, 10000)) +
  labs(
    x = "Age [years]",
    y = expression(bold("NfL [pg " * ml^-1 * "]"))) +
  theme_classic2() +
  theme(
  text = element_text(family = "Arial"),
    axis.line = element_line (linewidth = 1),
    axis.ticks = element_line (linewidth = 1),
    axis.text.x = element_text (size = 16, colour = "black"),
    axis.text.y = element_text (size = 16, colour = "black"),
    axis.title.x = element_text (size = 18, face = "bold"),
    axis.title.y = element_text (size = 18))

print(Figure_1B_plot)

ggsave("/Users/carinabergmann_1/Documents/01-Carina_PhD/28-NfL-Species-Paper/20-Resubmission/00-Figures/02-Fig1b.png", Figure_1B_plot, width = 8, height = 6, dpi = 300)


### Test for significance ### ONLY IF REQUESTED
#gam_model_mouse <- gam(NfL ~ s(Age_years, bs = "cs", k = 5), 
#                 data = Figure_1B[Figure_1B$Species == "Mouse", ])
#gam_model_cat <- gam(NfL ~ s(Age_years, bs = "cs", k = 5), 
#                 data = Figure_1B[Figure_1B$Species == "Cat", ])
#gam_model_dog <- gam(NfL ~ s(Age_years, bs = "cs", k = 5), 
#                 data = Figure_1B[Figure_1B$Species == "Dog", ])
#gam_model_horse <- gam(NfL ~ s(Age_years, bs = "cs", k = 5), 
#                 data = Figure_1B[Figure_1B$Species == "Horse", ])
#gam_model_human <- gam(NfL ~ s(Age_years, bs = "cs", k = 5), 
#                 data = Figure_1B[Figure_1B$Species == "Human", ])

#summary(gam_model_mouse)
#summary(gam_model_cat)
#summary(gam_model_dog)
#summary(gam_model_horse)
#summary(gam_model_human)

```

### *Figure 1 C* 
```{r}
# LOG PLOTS 
Linear_mouse <- lm(NfL_log ~ Age_years, data = NfL_mouse)
Linear_cat <- lm(NfL_log ~ Age_years, data = NfL_cat)
Linear_dog <- lm(NfL_log ~ Age_years, data = NfL_dog)
Linear_horse <- lm(NfL_log ~ Age_years, data = NfL_horse)
Linear_human <- lm(NfL_log ~ Age_years, data = NfL_human)

##### MOUSE #####
mouse_log <- ggplot(NfL_mouse, aes(x = Age_years, y = NfL_log)) +
  geom_point(color = "black",alpha = 0.5, position = position_jitter(width = 0.1, height = 0)) +
  geom_smooth(method = "lm", color = "blue", fill = "lightblue", alpha = 0.4) +  # Smooth curves
   xlim(0, 3) + ylim(0, 4) +
  labs(title = "", #log (NfL) as a function of age - Mouse
       x = "Age [years]",
    y = expression(bold("NfL [pg " * ml^-1 * "]"))) +
  theme_classic2() +
  theme(
  text = element_text(family = "Arial"),
    axis.line = element_line (linewidth = 1),
    axis.ticks = element_line (linewidth = 1),
    axis.text.x = element_text (size = 16, colour = "black"),
    axis.text.y = element_text (size = 16, colour = "black"),
    axis.title.x = element_text (size = 18, face = "bold"),
    axis.title.y = element_text (size = 18))
# Equation and R^2
summary_mouse <- summary(Linear_mouse)
coeff_mouse <- summary_mouse$coefficients
intercept_mouse <- round(coeff_mouse[1, 1], 3)
slope_mouse <- round(coeff_mouse[2, 1], 3)
r_squared_mouse <- round(summary_mouse$r.squared, 3)
eq_mouse <- paste0("y = ", slope_mouse, "x + ", intercept_mouse, 
                  "      \nR² = ", r_squared_mouse)

##### CAT ######
cat_log <- ggplot(NfL_cat, aes(x = Age_years, y = NfL_log)) +
  geom_point(color = "black",alpha = 0.5, position = position_jitter(width = 0.1, height = 0)) +
  geom_smooth(method = "lm", color = "blue", fill = "lightblue", alpha = 0.4) +  # Smooth curves
   xlim(0, 26) + ylim(-2, 4) + 
  labs(title = "", #log (NfL) as a function of age - Cat
        x = "Age [years]",
    y = expression(bold("NfL [pg " * ml^-1 * "]"))) + 
  theme_classic2() +
  theme(
  text = element_text(family = "Arial"),
    axis.line = element_line (linewidth = 1),
    axis.ticks = element_line (linewidth = 1),
    axis.text.x = element_text (size = 16, colour = "black"),
    axis.text.y = element_text (size = 16, colour = "black"),
    axis.title.x = element_text (size = 18, face = "bold"),
    axis.title.y = element_text (size = 18))
# Equation and R^2
summary_cat <- summary(Linear_cat)
coeff_cat <- summary_cat$coefficients
intercept_cat <- round(coeff_cat[1, 1], 3)
slope_cat <- round(coeff_cat[2, 1], 3)
r_squared_cat <- round(summary_cat$r.squared, 3)
eq_cat <- paste0("y = ", slope_cat, "x + ", intercept_cat, 
                  "      \nR² = ", r_squared_cat)

##### DOG #####
dog_log <- ggplot(NfL_dog, aes(x = Age_years, y = NfL_log)) +
  geom_point(color = "black",alpha = 0.5, position = position_jitter(width = 0.1, height = 0)) +
  geom_smooth(method = "lm", color = "blue", fill = "lightblue", alpha = 0.4) +  # Smooth curves
   xlim(0, 20) + ylim(-1, 4) + 
  labs(title = "", # log (NfL) as a function of age - Dog
        x = "Age [years]",
    y = expression(bold("NfL [pg " * ml^-1 * "]"))) +
  theme_classic2() +
  theme(
  text = element_text(family = "Arial"),
    axis.line = element_line (linewidth = 1),
    axis.ticks = element_line (linewidth = 1),
    axis.text.x = element_text (size = 16, colour = "black"),
    axis.text.y = element_text (size = 16, colour = "black"),
    axis.title.x = element_text (size = 18, face = "bold"),
    axis.title.y = element_text (size = 18))
# Equation and R^2
summary_dog <- summary(Linear_dog)
coeff_dog <- summary_dog$coefficients
intercept_dog <- round(coeff_dog[1, 1], 3)
slope_dog <- round(coeff_dog[2, 1], 3)
r_squared_dog <- round(summary_dog$r.squared, 3)
eq_dog <- paste0("y = ", slope_dog, "x + ", intercept_dog, 
                  "      \nR² = ", r_squared_dog)

##### HORSE #####
horse_log <- ggplot(NfL_horse, aes(x = Age_years, y = NfL_log)) +
  geom_point(color = "black",alpha = 0.5, position = position_jitter(width = 0.1, height = 0)) +
  geom_smooth(method = "lm", color = "blue", fill = "lightblue", alpha = 0.4) +  # Smooth curves
   xlim(0, 37) + ylim(0, 2) + 
  labs(title = "", #log (NfL) as a function of age - Horse
        x = "Age [years]",
    y = expression(bold("NfL [pg " * ml^-1 * "]"))) +
  theme_classic2() + 
  theme(
  text = element_text(family = "Arial"),
    axis.line = element_line (linewidth = 1),
    axis.ticks = element_line (linewidth = 1),
    axis.text.x = element_text (size = 16, colour = "black"),
    axis.text.y = element_text (size = 16, colour = "black"),
    axis.title.x = element_text (size = 18, face = "bold"),
    axis.title.y = element_text (size = 18)) 
# Equation and R^2
summary_horse <- summary(Linear_horse)
coeff_horse <- summary_horse$coefficients
intercept_horse <- round(coeff_horse[1, 1], 3)
slope_horse <- round(coeff_horse[2, 1], 3)
r_squared_horse <- round(summary_horse$r.squared, 3)
eq_horse <- paste0("y = ", slope_horse, "x + ", intercept_horse, 
                  "      \nR² = ", r_squared_horse)

##### HUMAN #####
human_log <- ggplot(NfL_human, aes(x = Age_years, y = NfL_log)) +
  geom_point(color = "black",alpha = 0.5, position = position_jitter(width = 0.1, height = 0)) +
  geom_smooth(method = "lm", color = "blue", fill = "lightblue", alpha = 0.4) +  # Smooth curves
   xlim(20, 110) + ylim(0, 3) + 
  labs(title = "", #log (NfL) as a function of age - Human
        x = "Age [years]",
    y = expression(bold("NfL [pg " * ml^-1 * "]"))) +
  theme_classic2() +
  theme(
  text = element_text(family = "Arial"),
    axis.line = element_line (linewidth = 1),
    axis.ticks = element_line (linewidth = 1),
    axis.text.x = element_text (size = 16, colour = "black"),
    axis.text.y = element_text (size = 16, colour = "black"),
    axis.title.x = element_text (size = 18, face = "bold"),
    axis.title.y = element_text (size = 18))
# Equation and R^2
summary_human <- summary(Linear_human)
coeff_human <- summary_human$coefficients
intercept_human <- round(coeff_human[1, 1], 3)
slope_human <- round(coeff_human[2, 1], 3)
r_squared_human <- round(summary_human$r.squared, 3)
eq_human <- paste0("y = ", slope_human, "x + ", intercept_human, 
                  "      \nR² = ", r_squared_human)


print (mouse_log)
eq_mouse

print (cat_log)
eq_cat

print (dog_log)
eq_dog

print (horse_log)
eq_horse

print (human_log)
eq_human


```

# **FIGURE 2**
```{r}
##### LOAD, MERGE AND PREPARE DATA #####
#Upload Excel file
Aging_study_summary_data <- read_excel("/Users/carinabergmann_1/Documents/01-Carina_PhD/28-NfL-Species-Paper/19-DATA/01_Bergmann_et_al_Figure2.xlsx")

#Calculate lifespan(days) 
Aging_study_summary_data$lifespan_days <- Aging_study_summary_data$Date_of_prep - Aging_study_summary_data$Date_of_birth

#Calculate the time to death from the time of blood collection 
Aging_study_summary_data$T1_ime_to_event <- Aging_study_summary_data$Age_at_death - Aging_study_summary_data$Age_BC1
Aging_study_summary_data$T2_ime_to_event <- Aging_study_summary_data$Age_at_death  - Aging_study_summary_data$Age_BC2
Aging_study_summary_data$T3_ime_to_event <- Aging_study_summary_data$Age_at_death  - Aging_study_summary_data$Age_BC3
Aging_study_summary_data$T4_ime_to_event <- Aging_study_summary_data$Age_at_death  - Aging_study_summary_data$Age_BC4
#Set animal status to 1 = dead
Aging_study_summary_data$status_os = 1

# Calculate random slope NfL
#Create one data file with all static factors
Aging_study_summary_data_static <- Aging_study_summary_data %>% 
  dplyr::select("Strain", "Animal_number","Batch", "Gender", "Earmark", "Date_dob" ="Date_of_birth","lifespan_days", "lifespan_months" = "Age_at_death", "T1_ime_to_event", "T2_ime_to_event", "T3_ime_to_event", "T4_ime_to_event","status_os")
#Create one long data file with all date data
Aging_study_summary_data_date_long <- Aging_study_summary_data %>%
  dplyr::select("Animal_number", "Date_1" = "Date_BC1","Date_2" =  "Date_BC2","Date_3" = "Date_BC3","Date_4" = "Date_BC4", "Date_dop" = "Date_of_prep") %>%
  pivot_longer(cols = starts_with("Date"), names_to = "Timepoint", values_to = "Date") %>%
  mutate(Timepoint = gsub("Date_", "", Timepoint))
#Create one long data file with all age data
Aging_study_summary_data_Age_long <- Aging_study_summary_data %>%
  dplyr::select("Animal_number", "Age_1" = "Age_BC1","Age_2" =  "Age_BC2","Age_3" = "Age_BC3","Age_4" = "Age_BC4", "Age_dop" = "Age_at_death") %>%
  pivot_longer(cols = starts_with("Age"), names_to = "Timepoint", values_to = "Age") %>%
  mutate(Timepoint = gsub("Age_", "", Timepoint))
#Create one long data file with all weight data
Aging_study_summary_data_Weight_long <- Aging_study_summary_data %>%
  dplyr::select("Animal_number", "Weight_1" = "Weight_BC1","Weight_2" =  "Weight_BC2","Weight_3" = "Weight_BC3","Weight_4" = "Weight_BC4", "Weight_dop" = "Weight_at_death") %>%
  pivot_longer(cols = starts_with("Weight"), names_to = "Timepoint", values_to = "Weight") %>%
  mutate(Timepoint = gsub("Weight_", "", Timepoint))
#Create one long data file with all NfL data
Aging_study_summary_data_NfL_long <- Aging_study_summary_data %>%
  dplyr:: select("Animal_number", "NfL_1" = "NfL_BC1","NfL_2" =  "NfL_BC2","NfL_3" = "NfL_BC3","NfL_4" = "NfL_BC4") %>%
  pivot_longer(cols = starts_with("NfL"), names_to = "Timepoint", values_to = "NfL") %>%
  mutate(Timepoint = gsub("NfL_", "", Timepoint))

#Put all long data files together in a new summary_long data file
Aging_study_summary_data_date_long <- full_join(Aging_study_summary_data_static, Aging_study_summary_data_date_long, by= "Animal_number")
Aging_study_summary_long <- full_join(Aging_study_summary_data_date_long, Aging_study_summary_data_Age_long,  by =c( "Animal_number", "Timepoint"))
Aging_study_summary_long <- full_join(Aging_study_summary_long, Aging_study_summary_data_Weight_long,  by =c( "Animal_number", "Timepoint"))
Aging_study_summary_long <- full_join(Aging_study_summary_long, Aging_study_summary_data_NfL_long,  by =c( "Animal_number", "Timepoint"))

Aging_study_summary_long <- Aging_study_summary_long %>%
  group_by(Animal_number) %>%
  mutate(lifespan_months = last(Age)) #creating a new variable

# CALCULATING RANDOM SLOPE
Aging_study_summary_long_nfl <- Aging_study_summary_long %>%
  drop_na(NfL) %>%
  group_by(Animal_number)

LMM_time_nfl_all <-(lmer(NfL  ~ Age + (1+Age|Animal_number), Aging_study_summary_long_nfl))
beta <- coef(LMM_time_nfl_all)$Animal_number
extracted1 <- tibble::rownames_to_column(beta, "Animal_number") %>% dplyr::select(-"(Intercept)") 
colnames(extracted1) <- c("Animal_number", "Random_slope_nfl")
Aging_study_summary_long <- merge(Aging_study_summary_long, extracted1, by = "Animal_number", all.y = TRUE)
## add optimizer so you don't get convergence error, " control = lmerControl(optimizer ="Nelder_Mead")"
LMM_time_nfl_all <-(lmer(NfL  ~ Age + (1+Age|Animal_number), data=Aging_study_summary_long_nfl, REML = TRUE, control = lmerControl(optimizer ="Nelder_Mead")))
summary(LMM_time_nfl_all) # gives back the results of LMM, with loaded lmerTest package it includes p-value
# takes first random_slope_nfl value per animal_number
#add random slope values to the Aging_study_summary_data 
Aging_study_summary_long_agg1 <- Aging_study_summary_long %>%
  group_by(Animal_number) %>%
  summarize(Random_slope_nfl = first(Random_slope_nfl))  

Aging_study_summary_data <- merge(Aging_study_summary_data, Aging_study_summary_long_agg1, by = "Animal_number")

# Calculate the last time point per animal_number --> Time_to_event
Aging_study_last_timepoint <- Aging_study_summary_long_nfl %>%
  group_by(Animal_number) %>%
  summarize(last_age = max(Age))   

Aging_study_summary_data <- merge(Aging_study_summary_data, Aging_study_last_timepoint, by = "Animal_number")

# Calculate Time to event for random slope 
Aging_study_summary_data$Time_to_event <- Aging_study_summary_data$Age_at_death - Aging_study_summary_data$last_age

```

### *Figure 2 A * 
```{r}
# CREATED WITH AFFINITY DESIGNER - WORKFLOW
```

### *Figure 2 B* 
```{r}
RS_beta <- coef(LMM_time_nfl_all)$Animal_number %>%
  tibble::rownames_to_column("Animal_number")

Aging_study_summary_long$Animal_number <- as.factor (Aging_study_summary_long$Animal_number)

# Plot
random_slope_plot <- ggplot(Aging_study_summary_long, aes(x = Age, y = NfL, color = Animal_number)) +
  geom_point(size = 2, alpha = 1, shape = 19, position = position_jitter(width = 0, height = 0), show.legend = FALSE) +
  geom_abline(
    data = RS_beta,
    aes(intercept = `(Intercept)`, slope = Age, color = Animal_number),
    show.legend = FALSE, alpha = 0.4) + 
  #scale_color_brewer(palette = "Set3") +
   xlim(18.9, 25) + ylim(0, 1500) +
  theme_classic2() +
  labs(title = "", # NfL vs. Age with individual random slopes
        x = "Age [months]",
    y = expression(bold("Plasma NfL [pg " * ml^-1 * "]"))) +
  theme_classic2() +
  theme(
  text = element_text(family = "Arial"),
    axis.line = element_line (linewidth = 1),
    axis.ticks = element_line (linewidth = 1),
    axis.text.x = element_text (size = 16, colour = "black"),
    axis.text.y = element_text (size = 16, colour = "black"),
    axis.title.x = element_text (size = 18, face = "bold"),
    axis.title.y = element_text (size = 18))
  theme(legend.position = "none") 

## Warning: Removed 63 rows containing missing values or values outside the scale range
# long format 220 rows --> 44 (date of prep) + 19 missing NfL @ time point 3/4 
  

print (random_slope_plot)

```

### *Figure 2 C* 
```{r}
#Aging_study_summary_data <- Aging_study_summary_data %>%
#  filter (Random_slope_nfl <= 150)

Linear_survival <- lm(Time_to_event ~ Random_slope_nfl, data = Aging_study_summary_data)

Aging_study_summary_data$Time_death <- Aging_study_summary_data$last_age + Aging_study_summary_data$Time_to_event


##### RANDOM SLOPE #####
survival_plot <- ggplot(Aging_study_summary_data, aes(x = Random_slope_nfl , y = Time_death)) +
  geom_point(color = "red",alpha = 0.5, position = position_jitter(width = 0, height = 0)) +
  geom_smooth(method = "lm", color = "blue", fill = "lightblue", alpha = 0.4) +  # Smooth curves
   xlim(0, 210) + ylim(17, 37) +
  labs(title = "", #NfL versus time until death based on last age [months]
       x = expression(bold("Random slope of NfL")),  y = "Time until death based on last age [months]") + 
    theme_classic2() +
  theme(
  text = element_text(family = "Arial"),
    axis.line = element_line (linewidth = 1),
    axis.ticks = element_line (linewidth = 1),
    axis.text.x = element_text (size = 16, colour = "black"),
    axis.text.y = element_text (size = 16, colour = "black"),
    axis.title.x = element_text (size = 18, face = "bold"),
    axis.title.y = element_text (size = 18),
  theme(legend.position = "none"))
  
# Equation and R^2
summary_survival <- summary(Linear_survival)
coeff_survival <- summary_survival$coefficients
intercept_survival <- round(coeff_survival[1, 1], 3)
slope_survival <- round(coeff_survival[2, 1], 3)
r_squared_survival <- round(summary_survival$r.squared, 3)
# p-value for slope
p_value_survival <- round(coeff_survival[2, 4], 4)  

eq_survival <- paste0("y = ", slope_survival, "x + ", intercept_survival, 
                  "      \nR² = ", r_squared_survival,
  "      \nP = ", p_value_survival)


eq_survival
print(survival_plot)


``` 

### *Figure 2 D* 
```{r, message = FALSE, warning = FALSE}
# Prepare data
Kaplan_Meier_mean <- Aging_study_summary_data %>%
  drop_na(Random_slope_nfl, Time_to_event, status_os) %>%
  mutate(NfL_cat2 = ifelse(Random_slope_nfl >= mean(Random_slope_nfl), "1", "0"))

mean(Kaplan_Meier_mean$Random_slope_nfl)
# Surv-object & Cox model
cox_model_mean <- coxph(Surv(Time_to_event, status_os) ~ NfL_cat2, data = Kaplan_Meier_mean)
summary_cox_mean <- summary(cox_model_mean)
# results
p_val_raw_mean <- summary_cox_mean$sctest["pvalue"]
hr_mean <- summary_cox_mean$conf.int[1, "exp(coef)"]
hr_low_mean <- summary_cox_mean$conf.int[1, "lower .95"]
hr_high_mean <- summary_cox_mean$conf.int[1, "upper .95"]
p_val_adj_mean <- min(p_val_raw_mean * 1, 1) # correct for multiple testing 
# Kaplan-Meier-Plot
plot_mean <- ggsurvplot(
  survfit(Surv(Time_to_event,status_os) ~ NfL_cat2, data = Kaplan_Meier_mean),
  data = Kaplan_Meier_mean,
  pval = TRUE,
  conf.int = TRUE,
  risk.table = TRUE,
  legend = "top",
  legend.labs = c("Low NfL","High NfL"),
  palette = c("darkseagreen", "brown2"), 
  ggtheme = theme_classic2(),
  xlab = "Time until death [months]", 
  ylab = "Survival probability",
  extend = TRUE)

plot_mean$plot <- plot_mean$plot +
  theme_classic2() +
  theme(
  text = element_text(family = "Arial"),
    axis.line = element_line (linewidth = 1),
    axis.ticks = element_line (linewidth = 1),
    axis.text.x = element_text (size = 16, color = "black"),
    axis.text.y = element_text (size = 16, color = "black"),
    axis.title.x = element_text (size = 18, face = "bold"),
    axis.title.y = element_text (size = 18,face = "bold"),
  legend.text = element_text(size = 14),
  legend.position = "top")

print(plot_mean)

```

### *Figure 2 E* 
```{r}
# Cox regression 
Aging_study_summary_data$Random_slope_nfl_z <- scale(Aging_study_summary_data$Random_slope_nfl)
Aging_study_summary_data$Gender <- as.factor(Aging_study_summary_data$Gender)

covariates <- c("Random_slope_nfl_z") 
#This means survival is modeled as a function of a covariate while adjusting for gender
univ_formulas <- sapply(covariates,
                        function(x) as.formula(paste('Surv(Time_to_event, status_os) ~ Random_slope_nfl_z +  last_age '))) # + Gender
# Fit Cox regression models for each covariate adjusted for gender
univ_models <- lapply(univ_formulas, function(x) { 
  tryCatch(coxph(x, data = Aging_study_summary_data), 
           error = function(e) NULL)})  
# tryCatch ensures that if an error occurs (e.g., missing data or convergence issues), the function returns NULL instead of crashing
# Extract results from each model
univ_results <- lapply(univ_models, 
                       function(x) { 
                         if (is.null(x)) return(NULL)
                         x_summary <- summary(x)
                         beta <- ifelse(!is.null(x_summary$coef),  
                                        signif(x_summary$coef[1], digits = 2), NA)
                         HR <- ifelse(!is.null(x_summary$coef), 
                                      signif(exp(x_summary$coef[1]), digits = 4), NA)
                         HR.confint.lower <- ifelse(!is.null(x_summary$conf.int), 
                                                    signif(x_summary$conf.int[,"lower .95"], digits = 4), NA)
                         HR.confint.upper <- ifelse(!is.null(x_summary$conf.int), 
                                                    signif(x_summary$conf.int[,"upper .95"], digits = 4), NA)
                         HR <- ifelse(!is.na(HR), 
                                      paste0(HR, " (", HR.confint.lower, "-", HR.confint.upper, ")"), NA)
                         # raw p-value
                         raw.p <- ifelse(!is.null(x_summary$coef), 
                                         x_summary$coef[1, "Pr(>|z|)"], NA)
                         # Bonferroni-> correct for 2 tests
                         adj.p <- ifelse(!is.na(raw.p), min(raw.p * 2, 1), NA)
                         p.value <- signif(adj.p, digits = 3)
                         wald.test <- ifelse(!is.null(x_summary$wald),
                                             signif(x_summary$wald["test"], digits = 2), NA)
                         res <- c(beta, HR, wald.test, p.value)
                         names(res) <- c("beta", "HR (95% CI for HR)", "wald.test", "p.value")
                         return(res)})

# Remove NULL results (models that failed)
univ_results_cleaned <- univ_results[!sapply(univ_results, is.null)]
# Converts the cleaned list into a transposed data frame (t()).
res <- t(as.data.frame(univ_results_cleaned, check.names = FALSE))
# Ensures the results are in a proper tabular format for visualization
as.data.frame(res)


# with gender --> p = 0.00718 
# without gender --> p = 0.00124
```



# **FIGURE 3**
```{r}
## DATA PLOTTED AND ANALYSED IN GRAPHPAD PRISM 10

# FIGURE 3 A #
NfL_young <- NfL_Species %>%
  filter (Sup_2_included == 1)
# FIGURE 3 B & C
NfL_young <-NfL_young[NfL_young$Conc_CV < 70 | is.na(NfL_young$Conc_CV), ]

Figure_3B <- NfL_young %>%
  group_by(Species)

Figure_3B <- subset(Figure_3B, !(Species %in% c("Cat", "Dog") & Age_years <= 1))
Figure_3B$NfL_weight <- Figure_3B$NfL * Figure_3B$'Average body weight'

Figure_3B <- Figure_3B %>%
  group_by(Species) %>%
  mutate(NfL_baseline = median(NfL, na.rm = TRUE)) %>%
  ungroup ()

Figure_3B$MLS_BW <- Figure_3B$`Maximum Age` + Figure_3B$`Average body weight`
Figure_3B$NfL_BW <- Figure_3B$NfL_baseline / Figure_3B$`Average body weight`

Correlation <- Figure_3B %>%
  group_by(Species) %>%
  summarise( 
    NfL_baseline = mean (NfL_baseline, na.rm =TRUE),
    AvBODY = mean (`Average body weight`, na.rm = TRUE),
    AvLIFE = mean (`Mean life expectancy`, na.rm = TRUE),
    MaxLIFE = mean (`Maximum Age`, na.rm = TRUE),
    MLS_BW = mean (MLS_BW, na.rm = TRUE),
    NfL_BW = mean (NfL_BW, na.rm = TRUE))

# SPEARMAN Correlations performed in PRISM for:
# NfL / BW ~ MaxAge
# NfL ~ BW + MaxAge
# MaxAge ~ BW
# NfL ~ BW 
# NfL ~ MaxAge 

```

# **FIGURE 4**
```{r}
# PLOT CREATED WITH (4a) GRAPHPAD PRISM 10 and (4b) EXCEL

Figure4_A <- NfL_Species %>%
  dplyr::filter (Species %in% c("Alpine ibex", "Barbary Sheep", "Bison", "Bonobo", "Bush dog", "Californian sea lion", "Camel", "Coati", "Collared peccary", "Dikdik", "Fallow deer", "Fennec", "Fur seal", "Goitered gazelle", "Giant panda", "Gorilla", "Grevy zebra",  "Kangaroo", "Kunekune pig", "Lama", "Lesser kudu", "Lion", "Lynx", "Markhor", "Meerkat", "Moose", "Okapi", "Orangutan", "Persian leopard", "Polar bear", "Reticulated giraffe", "Rhinoceros", "scimitar-horned oryx", "Seal", "Spider monkey", "Takin", "Tiger", "Wallaby", "Wapiti","Crocodylidae", "Lizard", "Snake", "Tortoise", "Bird")) %>%
  mutate(Species = dplyr::recode(Species,
                                 "scimitar-horned oryx" = "Sable antelope", "Goitered gazelle" = "Gazelle", "Crocodylidae" = "Crocodile")) %>%
 mutate(Species = factor(Species, levels = c("Alpine ibex", "Barbary Sheep", "Bison", "Bonobo", "Bush dog", "Californian sea lion", "Camel", "Coati", "Collared peccary", "Dikdik", "Fallow deer", "Fennec", "Fur seal", "Gazelle", "Giant panda", "Gorilla", "Grevy zebra",  "Kangaroo", "Kunekune pig", "Lama", "Lesser kudu", "Lion", "Lynx", "Markhor", "Meerkat", "Moose", "Okapi", "Orangutan", "Persian leopard", "Polar bear", "Reticulated giraffe", "Rhinoceros", "Sable antelope", "Seal", "Spider monkey", "Takin", "Tiger", "Wallaby", "Wapiti","Crocodile", "Lizard", "Snake", "Tortoise", "Bird")))

# Color
species_colors <- rep("black", length(levels(Figure4_A$Species)))
species_colors[(length(species_colors)-4):length(species_colors)] <- "red"

# Plot
ggplot(Figure4_A, aes(x = Species, y = NfL)) +
  geom_point(size = 1, color = "black", alpha = 0.6, 
           position = position_jitter(width = 0.1), inherit.aes = FALSE,
           aes(x = Species, y = NfL)) +
  labs(y = "NfL [pg/ml]", x = "") +
  scale_y_log10(
    limits = c(0.01, 100000),
    breaks = c(0.01, 0.1, 1, 10, 100, 1000, 10000, 100000),
    labels = scales::comma) +
  scale_x_discrete(labels = function(x) {
    mapply(function(label, color) {
      paste0("<span style='color:", color, "'>", label, "</span>")
    }, x, species_colors)}) +
  theme_classic2(base_size = 14) +
  theme(
    axis.text.x = element_markdown(angle = 45, hjust = 1))

```



# **Supplementary Table 1**
```{r}
S1_Table <- Figure_1_A %>%
  filter (Species %in% c("Mouse", "Cat", "Dog", "Human", "Horse")) %>%
  mutate(Species = factor (Species, levels = c("Mouse", "Cat", "Dog", "Human", "Horse"))) %>%
  mutate (Group = factor(Group, levels = c("Mouse <= 1 year", "Mouse > 1 - 2 years", "Mouse > 2 years", "Cat <= 7 years", "Cat > 7 - 14 years", "Cat > 14 years","Dog <= 6 years", "Dog > 6 - 12 years", "Dog > 12 years", "Human 20 - 40 years", "Human > 40 - 80 years", "Human > 80 years", "Horse <= 12.5 years", "Horse > 12.5 - 25 years", "Horse > 25 years"))) %>% 
  group_by (Species, Group) %>%
  summarise (
    Number = n(),
    Females = sum(Gender == "f", na.rm = TRUE),
    Males = sum(Gender == "m", na.rm = TRUE),
    Unknown_sex = sum (Gender == "u", na.rm = TRUE),
    Body_weight = mean (`Average body weight`, na.rm = TRUE),
    Max_age = mean (`Maximum Age`, na.rm = TRUE),
    Mean_life = mean (`Mean life expectancy`, na.rm = TRUE),
    Baseline_NfL = median(NfL, na.rm = TRUE))

S1_Table

```


# **Supplementary Table 2**
```{r}
NfL_young <- NfL_Species %>%
  filter (Sup_2_included == 1) %>%
  filter(!(Species %in% c("Cat", "Dog") & Age_years == 1)) 

NfL_young <- NfL_young %>%
  filter(is.na(Conc_CV) | !near(Conc_CV, 70.699229269)) # Remove one cattle  -->  Conc_CV of 70.7

S2_Table <- NfL_young %>%
   mutate(Species = factor (Species, levels = c("Alpaca", "Cat","Cattle", "Dog","Donkey", "Elephant", "Goat", "Horse", "Human", "Mouse", "Rabbit","Deer", "Sheep"))) %>%
  group_by(Species) %>%
  summarise (
    Number = n(),
    Females = sum(Gender == "f", na.rm = TRUE),
    Males = sum(Gender == "m", na.rm = TRUE),
    Unknown_sex = sum (Gender == "u", na.rm = TRUE),
    Body_weight = mean (`Average body weight`, na.rm = TRUE),
    Max_age = mean (`Maximum Age`, na.rm = TRUE),
    Mean_life = mean (`Mean life expectancy`, na.rm = TRUE),
    Baseline_NfL = median(NfL, na.rm = TRUE))


S2_Table


```


